Staged Abstract Interpreters

### 2019-08-14 paper/30_stageconc.tex
revision

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete interpreter
by changing the monad type and refactoring several primitive operations. We
begin by briefly introducing the Lightweight Modular Staging framework in Scala,
and then replay the same step in the unstaged counterpart. At last, we briefly
describe code generation.

@@

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the approach of
MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} that
uses syntactic quotation and escape, LMS distinguishes binding-time solely based
on types. LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary
type, indicating a value of @T@ will be known at the next stage. All operations
acting on @Rep[T]@ expressions will be residualized as generated code.

For the example we introduced in Section \ref{intro}, we have identified @b@ will 
be known later. To specialize it we write a function @snippet@ that provides a 
concrete value for @x@. After this, we may instantiate an @DslDriver@ object 
which provide the specialized program.

\begin{lstlisting}
  new DslDriver[Int, Int] {
    def power(b: Rep[Int], x: Int): Rep[Int] = if (x == 0) 1 else b * power(b, x-1)
    def snippet(b: Rep[Int]): Rep[Int] = power(b, 5)
  }
\end{lstlisting}

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as arrays and maps.
The idea is to construct intermediate representations (IR) for the next stage
program at the current stage. Such conversion from expressions of type @Rep[_]@ to
their IR is done by using the implicit design pattern. As we will see later,
implementing the staging support for custom classes is also straightforward.

@@

To implement the staged concrete interpreter, we reuse the exactly same concrete
components from Section \ref{unstaged_conc}, but we instantiate the binding-time
type @R[T]@ as @Rep[T]@.

\paragraph{Staged Monads} We also need to use monads that operate on staged
values. Note that the monadic objects such as @StateT@ or the internal value
@R[S] => M[(A,S)]@ are not staged, only the data @S@ or @A@ are staged.
This is why we can completely eliminate the overhead caused by the monadic layers.
Here we use the prefix name @Rep@ to differentiate unstaged monads and staged
monads, such as @RepReaderT@. So the @AnsM[_]@ is instantiated as the same monad
stack before, but with @Rep@ versions.

@@

@@

\begin{lstlisting}
  case class RepStateT[M[_]: RepMonad, S, A](run: Rep[S] => M[(A, S)]) {
    def flatMap[B](f: Rep[A] => StateT[M, S, B]): StateT[M, S, B] = ...
  }
\end{lstlisting}

@@

\paragraph{Code Generation}

@@

\begin{lstlisting}
  case class IRCompiledClo(f: Rep[((Value,Store)) => (Value, Store)], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[(Value, Store)]
\end{lstlisting}

>>>>

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to stage the concrete interpreter
by changing the monad type and refactoring several primitive operations. We
begin by briefly introducing the Lightweight Modular Staging framework in Scala,
and then replay the same steps from the unstaged counterpart. At last, we briefly
describe code generation.

@@

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} that
uses syntactic quotations and quasiquotations, LMS distinguishes binding-time
based on types. LMS provides a type constructor @Rep[T]@ where @T@ can be
an arbitrary type, indicating a value of @T@ will be known at the next stage.
All operations acting on @Rep[T]@ expressions will be residualized as generated
code.

A classic example for introducing multi-stage programming is the power function
that computes $b^x$, which is usually implemented as a recursive function:

\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

If @x@ is a value know at current stage, we may specialize the power function to
some value @x@ -- by unrolling the recursive calls. In LMS, this is fulfilled by
first adding the @Rep@ type to the variable known at next stage. In this case,
@b : Rep[Int]@ is know later, and @x : Int@ is know currently, as shown in the
below code recipe.
The way we use this @power@ function is to create a @DslDriver@ and override the
@snippet@ method give the currently know value @x@ (5 in this example).

\begin{lstlisting}
  new DslDriver[Int, Int] {
    def power(b: Rep[Int], x: Int): Rep[Int] = if (x == 0) 1 else b * power(b, x-1)
    def snippet(b: Rep[Int]): Rep[Int] = power(b, 5) // specialize the power to b^5
  }
\end{lstlisting}

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as lists and maps.
The idea behind the framework is to construct a sea-of-node intermediate
representations (IR) for the next stage program at the current stage
\cite{DBLP:conf/birthday/Rompf16}. For convinence, the conversion from
expressions of type @Rep[_]@ to their IR is done by using the implicit design
pattern. As we will see later, implementing staging and code generation
support for user-defined classes is also straightforward.

@@

To implement the staged concrete interpreter, we replay the steps from
instantiating unstaged concrete interpreter in Section \ref{unstaged_conc}.
But now we use the @Rep@ type to annotate value domains, environments and
stores, and redefine the staged version of monads and primitive operations.

\paragraph{Staged Monads}
We use the same structure of monad stack in the unstaged interpreter: a reader
monad with a state monad. But now the monads operate on staged values, for
brivity, we call them \textit{staged monads}. In the code, we also use a @Rep@
prefix on the constructors and types to differentiate them. But it is important
to note that objects of a case class @ReaderT@/@StateT@ are not staged, the
monadic computation like @R => M[A]@ are also not staged. Instead, the internal
data that these monads operate on are staged. The following code snippet shows
the idea (we use light gray to highlight what have been changed to staged
version except naming):

\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

The function @f@ passed to @flatMap@ is also a current-stage-known value, so
that all the invocation of @flatMap@s can be completed before generating code.
The fact that we only staged the data but not the monadic computation or monadic
values, is the reason that we can peel of the monad stack in the generated code.
Now the @AnsM[_]@ is instantiated as the same structure as before, but using the
@Rep@ versions.

@@

TODOTODO====

@@

@@

\subsection{A Little Bit Code Generation}

@@

\begin{lstlisting}
  type ValSt = (Value, Store)
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}
